---
title: "Draft - Final Paper"
author: "Shaya Engelman"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load necessary packages
library(tidyverse)
library(data.table)
library(mice)
library(GGally)
library(lubridate)

```

# Data Collection

## Read Datasets

```{r}
nyc_data <- fread("C:\\Users\\shaya\\OneDrive\\Documents\\repos\\Data698\\Research Project\\nyc_dataset.csv")
sf_data <- fread("C:\\Users\\shaya\\OneDrive\\Documents\\repos\\Data698\\Research Project\\sf_dataset.csv")
chicago_data <- fread("C:\\Users\\shaya\\OneDrive\\Documents\\repos\\Data698\\Research Project\\chicago_dataset.csv")
```

# Data Exploration

## NYC Data

```{r}
summary(nyc_data)
```

- Delete `total_length_miles`
- Handle missing values
- Check for outliers/anomalies
- Explore distributions of variables
- Identify relationships between variables
- Identify potential confounding variables

## Delete `total_length_miles`

```{r}
# Delete `total_length_miles`
nyc_data <- nyc_data %>%
  select(-total_length_miles)
```

## Handle Missing Values

Identify and address missing values in the dataset.


```{r}
# Handle missing values
missing_values <- nyc_data %>%
  summarise_all(~sum(is.na(.)))
print(missing_values)
```

### Remove Data from Before Citibike Introduction
Remove data from before the Citibike network was introduced, as it changes the dynamics of bike usage.

```{r}
# Delete rows from before Citibike was introduced: month 2013-06
nyc_data <- nyc_data %>%
  filter(month_year >= "2013-06-01")

# Verify missing values again
missing_values <- nyc_data %>%
  summarise_all(~sum(is.na(.)))
print(missing_values)
```

### Impute Missing Values for total_length_feet
Use the mice package to impute missing values for `total_length_feet`.

```{r}
# Impute missing values using mice
impute_data <- mice(nyc_data, m = 5, method = 'cart', seed = 123)
completed_data <- complete(impute_data)

# Update nyc_data with imputed values
nyc_data$total_length_feet <- completed_data$total_length_feet

# Verify missing values again
missing_values <- nyc_data %>%
  summarise_all(~sum(is.na(.)))
print(missing_values)
```

### Impute Missing Values for total_bike_counts and smoothed_volume
Impute missing values for total_bike_counts using citi_bike_count and for smoothed_volume using total_volume.

```{r}
# Impute missing values for `total_bike_counts` using `citi_bike_count`
model_bike_counts <- lm(total_bike_counts ~ citi_bike_count, data = nyc_data)
nyc_data <- nyc_data %>%
  mutate(total_bike_counts = ifelse(is.na(total_bike_counts), predict(model_bike_counts, newdata = nyc_data), total_bike_counts))

# Impute missing values for `smoothed_volume` using `total_volume`
model_smoothed_volume <- lm(smoothed_volume ~ total_volume, data = nyc_data)
nyc_data <- nyc_data %>%
  mutate(smoothed_volume = ifelse(is.na(smoothed_volume), predict(model_smoothed_volume, newdata = nyc_data), smoothed_volume))

# Verify missing values again
missing_values <- nyc_data %>%
  summarise_all(~sum(is.na(.)))

print(missing_values)
```

## Visualize the Data

```{r}
# Make line plots for all variables
nyc_data %>%
  gather(key = "variable", value = "value", -month_year) %>%
  ggplot(aes(x = month_year, y = value, color = variable)) +
  geom_line() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "NYC Data Over Time", x = "Month", y = "Value") +
  theme_minimal()
```

## Relationships Between Variables

```{r}
# Relationship between citi bike count and total bike counts to verify using bike share data as a proxy for bike usage using line plot on scaled data
nyc_data %>%
  select(month_year, citi_bike_count, total_bike_counts) %>%
  mutate(across(c(citi_bike_count, total_bike_counts), scale)) %>%
  rename("Citi Bike Usage" = citi_bike_count, "Total Bike Usage" = total_bike_counts) %>%
  melt(id.vars = "month_year") %>%
  ggplot(aes(x = month_year, y = value, color = variable)) +
  geom_line(size = 1.5, alpha = 0.8) + 
  labs(title = "Bike Usage Trends For All Bikes and Citi Bike", x = "Month", y = "Scaled Value") + 
  theme_minimal() +
  theme(
    legend.position = c(0.85, 0.1),  # Position the legend inside the plot
    plot.title = element_text(size = 35),  # Increase title size
    legend.title = element_blank(),  # Remove legend title
    legend.text = element_text(size = 15)  # Increase legend text size
  ) +
  scale_color_brewer(palette = "Set1")  # Use a different color palette
```

```{r}
# Select the relevant columns and compute the correlation matrix
cor_data <- nyc_data %>% 
  select(smoothed_volume, citi_bike_count, Total_Injured, total_length_feet, biking_proportion)

# Create the correlation plot using ggcorplot
ggcorr(cor_data,
       label = TRUE,
       label_round = 2,
       label_size = 4,
       colors = c("blue", "white", "red"),
       title = "Correlation Matrix Heatmap") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 45, vjust = 1)
  )

# Use cor_data for a matrix
cor(cor_data, use = "pairwise.complete.obs")
```


```{r}
# Create the new dataframe with the selected columns and rename them
selected_data_nyc <- nyc_data %>%
  select(
    date = month_year,
    mv_traffic = smoothed_volume,
    bike_lane_addition = total_length_feet,
    bike_count = citi_bike_count,
    accidents_count = Total_Injured,
    total_killed = Total_Killed
  )

# Display the first few rows of the new dataframe
head(selected_data_nyc)
```

```{r}
# Create a column with the ratio of bike_count to mv_traffic
selected_data_nyc <- selected_data_nyc %>%
  mutate(bike_to_traffic_ratio = bike_count / mv_traffic)

# Create a column with cumulative sum of bike_lane_addition
selected_data_nyc <- selected_data_nyc %>%
  mutate(cumulative_bike_lane = cumsum(bike_lane_addition))

# Display the first few rows of the updated dataframe
head(selected_data_nyc)
```

```{r}
# Plot the cumulative sum of bike lane additions against bike_to_traffic_ratio
ggplot(selected_data_nyc, aes(x = cumulative_bike_lane, y = bike_to_traffic_ratio)) +
  geom_point() +
  labs(title = "Cumulative Bike Lane Additions vs. Bike to Traffic Ratio", x = "Cumulative Bike Lane Additions", y = "Bike to Traffic Ratio") +
  theme_minimal()
```

```{r}
# Decompose the time series of bike_count and mv_traffic and recreate the bike_to_traffic_ratio with the decomposed values

# First, convert the bike count and motor vehicle traffic to time series objects

selected_data_nyc$bike_count <- ts(selected_data_nyc$bike_count, frequency = 12, start = c(2013, 6))
selected_data_nyc$mv_traffic <- ts(selected_data_nyc$mv_traffic, frequency = 12, start = c(2013, 6))

decomposed_bike_count <- stl(selected_data_nyc$bike_count, s.window = "periodic")
decomposed_mv_traffic <- stl(selected_data_nyc$mv_traffic, s.window = "periodic")

# Display the first few rows of the updated dataframe
head(selected_data_nyc)
```

```{r}
# Plot the bike_to_traffic_ratio over time
ggplot(selected_data_nyc, aes(x = date, y = bike_to_traffic_ratio)) +
  geom_line() +
  labs(title = "Decomposed Bike to Traffic Ratio Over Time", x = "Date", y = "Bike to Traffic Ratio") +
  theme_minimal()
``` 

```{r}  
# Convert bike_to_traffic_ratio to a time series and decompose it
selected_data_nyc$bike_to_traffic_ratio <- ts(selected_data_nyc$bike_to_traffic_ratio, frequency = 12, start = c(2013, 6))
decomposed_ratio <- stl(selected_data_nyc$bike_to_traffic_ratio, s.window = "periodic")

# Plot the decomposed bike_to_traffic_ratio
autoplot(decomposed_ratio) + ggtitle("Decomposed Bike to Traffic Ratio") +
  theme_minimal()

# Extract the trend and seasonality components
trend_nyc <- decomposed_ratio$time.series[, "trend"]
seasonal_nyc <- decomposed_ratio$time.series[, "seasonal"]
```

```{r}
# Plot the trend of bike_to_traffic_ratio against the trend of bike_lane_addition
ggplot(selected_data_nyc, aes(x = trend_nyc, y = cumulative_bike_lane)) +
  geom_point() +
  labs(title = "Trend of Bike to Traffic Ratio vs. Trend of Cumulative Bike Lane Additions", x = "Trend of Bike to Traffic Ratio", y = "Trend of Cumulative Bike Lane Additions") +
  theme_minimal()
``` 

```{r}
model_linear <- lm(cumulative_bike_lane ~ trend_nyc, data = selected_data_nyc)
summary(model_linear)

model_log <- lm(cumulative_bike_lane ~ log(trend_nyc + 1), data = selected_data_nyc)
summary(model_log)

model_exp <- lm(log(cumulative_bike_lane + 1) ~ trend_nyc, data = selected_data_nyc)
summary(model_exp)

model_logistic <- nls(cumulative_bike_lane ~ SSlogis(trend_nyc, Asym, xmid, scal), data = selected_data_nyc)
summary(model_logistic)

ggplot(selected_data_nyc, aes(x = trend_nyc, y = cumulative_bike_lane)) +
  geom_point() +
  #geom_smooth(method = "lm", formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", formula = y ~ log(x + 1), se = FALSE, col = "red") + 
  #geom_smooth(method = "lm", formula = y ~ exp(x), se = FALSE, col = "green") + 
  geom_line(aes(y = fitted(model_logistic)), col = "purple") +
  labs(title = "Model Fits: Logarithmic, Logistic",
       x = "Trend of Bike to Traffic Ratio",
       y = "Cumulative Bike Lane Additions") +
  theme_minimal() +
  #scale_color_manual(values = c("blue", "red", "green", "purple"),
  #                   labels = c("Linear", "Log", "Exponential", "Logistic")) +
  theme(legend.position = "bottom")
``` 

```{r}
ggplot(selected_data_nyc, aes(x = trend, y = cumulative_bike_lane)) +
  geom_point(aes(color = "Actual Ratio")) +
  geom_smooth(method = "lm", formula = y ~ log(x + 1), se = FALSE, aes(color = "Logarithmic")) + 
  geom_line(aes(y = fitted(model_logistic), color = "Logistic")) +
  labs(title = "NYC Model Fits: Logarithmic, Logistic",
       y = "Total Amount of Bike Lanes",
       x = "Trend of Bike to Traffic Ratio") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red", "purple"),
                     labels = c("Actual Ratio", "Logarithmic", "Logistic")) +
  theme(
    legend.position = c(0.78, 0.17),
    plot.title = element_text(size = 30),
    legend.title = element_blank(),
    legend.text = element_text(size = 25),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  )
```

The logistic model seems to be the best fit for the data. The pattern of the trend of cumulative bike lane additions against the trend of bike to traffic ratio is similar to a logistic curve, indicating that there may be a saturation point for bike lane additions beyond which the bike to traffic ratio does not increase significantly. However, while a typical logistic curve has a carrying capacity, after which the curve completely flattens out, the data seems to retain a very slight upward trend even after the inflection point. This suggests that the relationship between bike lane additions and the bike to traffic ratio may be more complex than a simple logistic curve can capture. We will keep this as a potential model but explore adding more complexity as well. Additionally, we will consider a logarithmic model as it fits the model reasonable well and also keeps the positive trend after the inflection point.

```{r}
# Create a new column with the predicted values from the logistic model
selected_data_nyc$predicted_logistic <- predict(model_logistic)

# Create a new column with the predicted values from the logarithmic model
selected_data_nyc$predicted_log <- predict(model_log)
```

# San Francisco Data

```{r}
# Display the first few rows of the San Francisco dataset
head(sf_data)
```

# Repeat a similar process for the San Francisco data

```{r}
# Delete `total_length_miles`
sf_data <- sf_data %>%
  select(-total_length_miles)

# Handle missing values
missing_values <- sf_data %>%
  summarise_all(~sum(is.na(.)))
print(missing_values)
```

```{r}
# Remove data from before the bike share network was introduced
sf_data <- sf_data %>%
  filter(Month >= "2017-06-")

# Verify missing values again
missing_values <- sf_data %>%
  summarise_all(~sum(is.na(.)))
print(missing_values)
```

```{r}
# Print the rows with missing values in any column or a 0 in bay_wheels_count
sf_data_filtered <- sf_data_complete %>%
  filter(if_any(everything(), is.na) | bay_wheels_count == 0)

print(sf_data_filtered)
```  

```{r}
# Convert Month column to Date format
sf_data <- sf_data %>%
  mutate(Month = ymd(paste0(Month, "-01")))

# Create a complete sequence of months
full_months <- data.frame(Month = seq(min(sf_data$Month, na.rm = TRUE), max(sf_data$Month, na.rm = TRUE), by = "month"))

# Merge with your original data
sf_data_complete <- full_months %>%
  left_join(sf_data, by = "Month")

# View the data
head(sf_data_complete)
```

```{r}
# Summarize missing data
#summary(sf_data_complete)

# Impute missing values
imputed_data <- mice(sf_data_complete, m = 5, method = 'cart', seed = 123)

# Extract the completed dataset
sf_data_imputed <- complete(imputed_data, 1)

# View the imputed data
head(sf_data_imputed)
```

```{r}
# verify missing values again
missing_values <- sf_data_imputed %>%
  summarise_all(~sum(is.na(.)))
print(missing_values)
```

```{r}
# Create a new dataframe with the selected columns and rename them
selected_data_sf <- sf_data_imputed %>%
  select(
    date = Month,
    mv_traffic = Total_Volume_ADT,
    bike_lane_addition = total_length_feet,
    bike_count = bay_wheels_count,
    accidents_count = Total_Injured,
    total_killed = Total_Killed
  )

# Display the first few rows of the new dataframe
head(selected_data_sf)
```

```{r}
# Create a column with the ratio of bike_count to mv_traffic
selected_data_sf <- selected_data_sf %>%
  mutate(bike_to_traffic_ratio = bike_count / mv_traffic)

# Create a column with cumulative sum of bike_lane_addition
selected_data_sf <- selected_data_sf %>%
  mutate(cumulative_bike_lane = cumsum(bike_lane_addition))

# Display the first few rows of the updated dataframe
head(selected_data_sf)
``` 

```{r}
# Plot the cumulative sum of bike lane additions over time
ggplot(selected_data_sf, aes(x = date, y = cumulative_bike_lane)) +
  geom_line() +
  labs(title = "Cumulative Bike Lane Additions Over Time", x = "Date", y = "Cumulative Bike Lane Additions") +
  theme_minimal()
```    

```{r}
# Plot the cumulative sum of bike lane additions against bike_to_traffic_ratio
ggplot(selected_data_sf, aes(x = cumulative_bike_lane, y = bike_to_traffic_ratio)) +
  geom_point() +
  labs(title = "Cumulative Bike Lane Additions vs. Bike to Traffic Ratio", x = "Cumulative Bike Lane Additions", y = "Bike to Traffic Ratio") +
  theme_minimal()
```   

```{r}
# Decompose the time series of bike_count and mv_traffic and recreate the bike_to_traffic_ratio with the decomposed values

# First, convert the bike count and motor vehicle traffic to time series objects
selected_data_sf$bike_count <- ts(selected_data_sf$bike_count, frequency = 12, start = c(2017, 6))
selected_data_sf$mv_traffic <- ts(selected_data_sf$mv_traffic, frequency = 12, start = c(2017, 6))

decomposed_bike_count_sf <- stl(selected_data_sf$bike_count, s.window = "periodic")
decomposed_mv_traffic_sf <- stl(selected_data_sf$mv_traffic, s.window = "periodic")
```

```{r}
# Convert bike_to_traffic_ratio to a time series and decompose it
selected_data_sf$bike_to_traffic_ratio <- ts(selected_data_sf$bike_to_traffic_ratio, frequency = 12, start = c(2017, 6))
decomposed_ratio_sf <- stl(selected_data_sf$bike_to_traffic_ratio, s.window = "periodic")

# Plot the decomposed bike_to_traffic_ratio
autoplot(decomposed_ratio_sf) + ggtitle("Decomposed Bike to Traffic Ratio") +
  theme_minimal()
```

```{r}
# Extract the trend and seasonality components
trend_sf <- decomposed_ratio_sf$time.series[, "trend"]
seasonal_sf <- decomposed_ratio_sf$time.series[, "seasonal"]

# Plot the trend of bike_to_traffic_ratio against the trend of bike_lane_addition
ggplot(selected_data_sf, aes(x = trend_sf, y = cumulative_bike_lane)) +
  geom_point() +
  labs(title = "Trend of Bike to Traffic Ratio vs. Trend of Cumulative Bike Lane Additions", x = "Trend of Bike to Traffic Ratio", y = "Trend of Cumulative Bike Lane Additions") +
  theme_minimal()
```  

```{r}
# Plot the NYC and SF trend of bike_to_traffic_ratio against the trend of bike_lane_addition on separate plots in the same figure
ggplot(selected_data_nyc, aes(x = trend_nyc, y = cumulative_bike_lane)) +
  geom_point() +
  labs(title = "NYC: Trend of Bike to Traffic Ratio vs. Trend of Cumulative Bike Lane Additions", x = "Trend of Bike to Traffic Ratio", y = "Trend of Cumulative Bike Lane Additions") +
  theme_minimal() +
  facet_wrap(~ "NYC") +
  theme(plot.title = element_text(hjust = 0.5))
```   

```{r}
# Install and load necessary packages
library(gridExtra)

# Create the individual plots
plot_sf <- ggplot(selected_data_sf, aes(x = trend_sf, y = cumulative_bike_lane)) + 
  geom_point() +
  labs(title = "SF: Trend of Bike to Traffic Ratio vs. Trend of Cumulative Bike Lane Additions",
       x = "Trend of Bike to Traffic Ratio",
       y = "Trend of Cumulative Bike Lane Additions") +
  theme_minimal()

plot_nyc <- ggplot(selected_data_nyc, aes(x = trend_nyc, y = cumulative_bike_lane)) + 
  geom_point() +
  labs(title = "NYC: Trend of Bike to Traffic Ratio vs. Trend of Cumulative Bike Lane Additions",
       x = "Trend of Bike to Traffic Ratio",
       y = "Trend of Cumulative Bike Lane Additions") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Arrange plots side by side
grid.arrange(plot_sf, plot_nyc, ncol = 2)
```

```{r}
model_linear_sf <- lm(cumulative_bike_lane ~ trend_sf, data = selected_data_sf)
summary(model_linear_sf)

model_log_sf <- lm(cumulative_bike_lane ~ log(trend_sf + 1), data = selected_data_sf)
summary(model_log_sf)

model_exp_sf <- lm(log(cumulative_bike_lane + 1) ~ trend_sf, data = selected_data_sf)
summary(model_exp_sf)

model_logistic_sf <- nls(cumulative_bike_lane ~ SSlogis(trend_sf, Asym, xmid, scal),
                         data = selected_data_sf,
                          start = list(Asym = max(selected_data_sf$cumulative_bike_lane) * 1.1,
                                       xmid = 4,
                                       scal = 0.5))
summary(model_logistic_sf)
```

```{r}
ggplot(selected_data_sf, aes(x = trend_sf, y = cumulative_bike_lane)) +
  geom_point(aes(color = "Actual Ratio")) +
  geom_smooth(method = "lm", formula = y ~ log(x + 1), se = FALSE, aes(color = "Logarithmic")) +
  geom_line(aes(y = fitted(model_logistic_sf), color = "Logistic")) +
  labs(title = "SF Model Fits: Logarithmic, Logistic",
       x = "Trend of Bike to Traffic Ratio",
       y = "Trend of Cumulative Bike Lane Additions") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red", "purple"),
                     labels = c("Actual Ratio", "Logarithmic", "Logistic")) +
  theme(
    legend.position = c(0.78, 0.17),
    plot.title = element_text(size = 30),
    legend.title = element_blank(),
    legend.text = element_text(size = 25),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  )
```

The San Fracisco data is a smaller dataset than the NYC data, but it shows a similar trend in the relationship between bike lane additions and the bike to traffic ratio. It also seems to follow a logistic curve pattern, indicating a saturation point for bike lane additions. However, this too fits a logarthimic model reasonably well and retains a slight positive trend after the inflection point. We will keep the logistic model as a potential model but explore adding more complexity as well. Additionally, we will consider a logarithmic model as it fits the model reasonably well and also keeps the positive trend after the inflection point. The logarithmic model is also a more logical fit for bike lane additions, as it is unlikely for these to be a strict saturation point where no more bike lanes can be added, while the effect slowing down as more bike lanes are added is more likely.

## Analyze total accidetnts and total killed relative to bike to traffic ratio

```{r}
# Create a new column with the combined injuries and deaths
nyc_data_temp <- nyc_data %>%
  mutate(total_injuries_deaths = Total_Injured + 5*Total_Killed)

# Plot the trend of bike_to_traffic_ratio against the trend of total_injuries_deaths
ggplot(nyc_data_temp, aes(x = trend_nyc, y = total_injuries_deaths)) +
  geom_point() +
  labs(title = "NYC: Trend of Bike to Traffic Ratio vs. Trend of Total Injuries and Deaths", x = "Trend of Bike to Traffic Ratio", y = "Trend of Total Injuries and Deaths") +
  theme_minimal()
```  

```{r}
nyc_data_temp <- nyc_data_temp %>%
  rename(date = month_year)

# Now you can perform the join
selected_data_nyc <- selected_data_nyc %>%
  left_join(nyc_data_temp, by = "date")
```  

```{r}
# Plot total_injuries_deaths against cumulative_bike_lane
ggplot(selected_data_nyc, aes(x = cumulative_bike_lane, y = total_injuries_deaths)) +
  geom_point() +
  labs(title = "NYC: Cumulative Bike Lane Additions vs. Total Injuries and Deaths", x = "Cumulative Bike Lane Additions", y = "Total Injuries and Deaths") +
  theme_minimal()
``` 

```{r}
# Plot total_injuries_deaths against traffic volume
ggplot(selected_data_nyc, aes(x = mv_traffic, y = total_injuries_deaths)) +
  geom_point() +
  labs(title = "NYC: Motor Vehicle Traffic Volume vs. Total Injuries and Deaths", x = "Motor Vehicle Traffic Volume", y = "Total Injuries and Deaths") +
  theme_minimal()
```  

```{r}
# Try a linear regression model for the relationship between total_injuries_deaths and traffic volume
model_linear_nyc_injuries <- lm(total_injuries_deaths ~ mv_traffic, data = selected_data_nyc)
summary(model_linear_nyc_injuries)
```

While there does not seem to be a statistically significant relationship between the bike to traffic ratio and the total injuries and deaths, there is a clear positive relationship between the motor vehicle traffic volume and the total injuries and deaths. That being the case, there must be a relationship between people choosing fewer cars (indicated by a higher bike to traffic ratio) and fewer injuries and deaths. This wasn't captured in the linear model, but that is likely due to it beinng a very small effect relative to total traffic volume. Having established this relationship, and also a positive logarithmic relationship between bike lane additions and the bike to traffic ratio, we can infer a positive relationship between bike lane additions and fewer injuries and deaths.

```{r}
# Model logarithmic relationship between bike lane additions and bike to traffic ratio, linear relationship between bike to traffic ratio and total injuries and deaths, linear relationship between motor vehicle traffic volume and total injuries and deaths

log_lane_bike_ratio <- lm(cumulative_bike_lane ~ log(trend_nyc + 1), data = selected_data_nyc)
linear_bike_ratio_injuries <- lm(total_injuries_deaths ~ trend_nyc, data = selected_data_nyc)
linear_traffic_injuries <- lm(total_injuries_deaths ~ mv_traffic, data = selected_data_nyc)

# Print the summary of the models
summary(log_lane_bike_ratio)
summary(linear_bike_ratio_injuries)
summary(linear_traffic_injuries)

# Tidy up the model summaries and remove intercept rows
tidy_log_lane_bike_ratio <- tidy(log_lane_bike_ratio) %>% filter(term != "(Intercept)")
tidy_bike_ratio_injuries <- tidy(bike_ratio_injuries) %>% filter(term != "(Intercept)")
tidy_traffic_injuries <- tidy(traffic_injuries) %>% filter(term != "(Intercept)")

# Rename the terms for clarity
tidy_log_lane_bike_ratio <- tidy_log_lane_bike_ratio %>% mutate(term = "Log-Transformed Bike-to-Traffic Ratio")
tidy_bike_ratio_injuries <- tidy_bike_ratio_injuries %>% mutate(term = "Bike-to-Traffic Ratio")
tidy_traffic_injuries <- tidy_traffic_injuries %>% mutate(term = "Motor Vehicle Traffic Volume")

# Get the model statistics
glance_log_lane_bike_ratio <- glance(log_lane_bike_ratio)
glance_bike_ratio_injuries <- glance(bike_ratio_injuries)
glance_traffic_injuries <- glance(traffic_injuries)

# Combine the model statistics with the summaries
model_summaries <- bind_rows(
  mutate(tidy_log_lane_bike_ratio, Model = "Log Lane-Bike Ratio Model", `R-Squared` = glance_log_lane_bike_ratio$r.squared),
  mutate(tidy_bike_ratio_injuries, Model = "Bike Ratio-Injuries Model", `R-Squared` = glance_bike_ratio_injuries$r.squared),
  mutate(tidy_traffic_injuries, Model = "Traffic-Injuries Model", `R-Squared` = glance_traffic_injuries$r.squared)
)

# Color insignificant p-values red
model_summaries <- model_summaries %>%
  mutate(`P-Value` = ifelse(p.value > 0.05, cell_spec(p.value, color = "red"), p.value)) %>%
  select(Model, Term = term, Estimate = estimate, `Standard Error` = std.error, `t-Value` = statistic, `P-Value`, `R-Squared`)

# Create and format the table
model_summaries %>%
  kable("html", escape = FALSE, caption = "Model Summaries for Various Relationships") %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = TRUE, background = "#4CAF50", color = "white") %>%
  row_spec(seq(1, nrow(model_summaries), by = 2), background = "#f2f2f2") %>%
  row_spec(seq(2, nrow(model_summaries), by = 2), background = "#e0e0e0") %>%
  column_spec(1:7, border_left = TRUE, border_right = TRUE) %>%
  add_header_above(c(" " = 1, "Model Summary" = 6))

```

```{r}
# Repeat for San Francisco data
# Calculate cumulative bike lane additions and total injuries and deaths
selected_data_sf <- selected_data_sf %>%
  mutate(cumulative_bike_lane = cumsum(bike_lane_addition),
          total_injuries_deaths = accidents_count + 5*total_killed)

# Plot traffic volume against total injuries and deaths
ggplot(selected_data_sf, aes(x = mv_traffic, y = total_injuries_deaths)) +
  geom_point() +
  labs(title = "SF: Motor Vehicle Traffic Volume vs. Total Injuries and Deaths", x = "Motor Vehicle Traffic Volume", y = "Total Injuries and Deaths") +
  theme_minimal()
```  

```{r}
# Model logarithmic relationship between bike lane additions and bike to traffic ratio, linear relationship between bike to traffic ratio and total injuries and deaths, linear relationship between motor vehicle traffic volume and total injuries and deaths

log_lane_bike_ratio_sf <- lm(cumulative_bike_lane ~ log(trend_sf + 1), data = selected_data_sf)
linear_bike_ratio_injuries_sf <- lm(total_injuries_deaths ~ trend_sf, data = selected_data_sf)
linear_traffic_injuries_sf <- lm(total_injuries_deaths ~ mv_traffic, data = selected_data_sf)

# Print the summary of the models
summary(log_lane_bike_ratio_sf)
summary(linear_bike_ratio_injuries_sf)
summary(linear_traffic_injuries_sf)
```